---
title: "Iris Dataset Analysis - Change 2"
format: pdf
editor: visual
---

# 1 - Introduction

We use the built-in `iris` dataset within `R` to demonstrate a simple data visualisation. This dataset is comprised of 150 observations, each recording four characteristics of the iris flowers (petal length, petal width, sepal length, sepal width) and their species.

## *Data Loading*

Lets bring the data into our local environment and view the first few rows of the dataset.

```{r, echo = FALSE}
data(iris)
head(iris)
```

## *Data Structure*

Diving deeper into the datasets composition and the various data types.

```{r, echo = FALSE}
str(iris)
```

# 2 - Visualisation

## *Violin Plots for petal and sepal measurements across each species*

These provide information on both the distribution and the frequency of values which can describe the differences between the species. **`ggplot2`** is used to plot these.

```{r, echo = FALSE,message=FALSE,warning=FALSE}
library(ggplot2)
# violin plot for sepal length across all species
ggplot(iris, aes(x = Species, y = Sepal.Length, fill = Species)) +
  geom_violin() +
  labs(x = "Species", y = "Sepal Length (cm)", title = "Violin Plot for Sepal Length") +
  theme_bw()

# violin plot for sepal width 
ggplot(iris, aes(x = Species, y = Sepal.Width, fill = Species)) +
  geom_violin() +
  labs(x = "Species", y = "Sepal Width (cm)", title = "Violin Plot for Sepal Width") +
  theme_bw()

# violin plot for petal length
ggplot(iris, aes(x = Species, y = Petal.Length, fill = Species)) +
  geom_violin() +
  labs(x = "Species", y = "Petal Length (cm)", title = "Violin Plot for Petal Length") +
  theme_bw()

# violin plot for petal width
ggplot(iris, aes(x = Species, y = Petal.Width, fill = Species)) +
  geom_violin() +
  labs(x = "Species", y = "Petal Width (cm)", title = "Violin Plot for Petal Width") +
  theme_bw()
```

## *Visualisations Summary*

These plots clearly show the distinctive differences among the 3 species across both petal and sepal lengths and widths Setosa flowers have a tight distribution in their petal length and width, around 1.5 and 0.25 respectively, whereas both versicolor and virginica have much longer lengths in this category with a wider, more spread out distribution.
